Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-01-14 Manuscript/Manuscript.tex

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

% Pandoc syntax highlighting

% Pandoc citation processing

@@

\sout{Research into any field with a hierarchical or clustered nature of
observations may yield multilevel data. In the typical case, individuals
are nested within groups, but there are many different types of
multilevel data. In the medical field, clustering occurs at e.g., the
hospital or center level in registry data, or at the study-level in
meta-analyses (IPDMA). In the social sciences and official statistics we
can find clustering e.g.~at the country-level, or as imposed by a
multi-stage sampling design. For the sake of legibility, we will refer
to the grouping variable as `cluster', and the grouped variable as
`(sample) unit' throughout this paper. And, for reasons of brevity, we
only discuss clustering between units, not within units (e.g.,
timeseries or longitudinal data). Also, add that we'll only discuss two
levels, not more?}

In many contemporary data analysis efforts, some form of hierarchical or
clustered structure is recorded. Ignoring such structures may be harmful
to the inferences and can yield biased results. On the other hand,
analyzing such multilevel data requires specialized techniques that take
the clustered structure into account. Imagine a case where cross-level
interactions between unit-level variables and cluster-level variables
are present. The cluster to which a unit belongs may then influence the
unit-level observations - and vice versa - for each of the units that
make up the cluster \citep{hox17}. These relations can and should be
taken into account when developing analysis models for multilevel data
for the simple reason that groups of observations share some common
variance.

@@

The process of analyzing multilevel data is further complicated when not
all observations are observed. Just as with single level data,
missingness may occur at the unit level. But with multiple levels of
data comes the potential for clustered missingness. Missingness in
multilevel data can therefore be categorized into two general patterns:
systematic missingness and sporadic missingness, see \citet{resc13}. We
have visualized the difference between these types of missingness in
Figure 1. The figure shows an \(n \times p\) set
\(\mathbf{X} = X_1, \dots, X_p\), with \(n=5\) units and \(p=3\) columns
distributed over 3 clusters. Column \texttt{X1} is completely observed.
Column \texttt{X2} is systematically missing and column \texttt{X3} is
sporadically missing.\footnote{Explain why. GV: niet nodig}

@@

\[
\begin{matrix} 
                          &X1     & X2          & X3            \\
   \hline
   \mathrm{cluster 1}\{ \ &x_{11} &x_{12}       &\mathtt{NA}    \\
                             \ &x_{21}  &x_{22}       &x_{32}         \\
   \mathrm{cluster 2}\{  &x_{31} &\mathtt{NA}  &x_{33}         \\
                          &x_{41} &\mathtt{NA}  &\mathtt{NA}    \\
   \mathrm{cluster 3}\{  &x_{51} &x_{52}       &x_{53}         \\
\end{matrix} 
\] {[}GV Ik vind de latex tabel niet tof. {]}

@@

>>>>

@@

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

In many contemporary data analysis efforts, some form of hierarchical or
clustered structure is recorded.\footnote{Add IPDMA back in: patient
  level data from multiple studies.} Ignoring such structures may be
harmful to the inferences and can yield biased results. On the other
hand, analyzing such multilevel data requires specialized techniques
that take the clustered structure into account. Imagine a case where
cross-level interactions between unit-level variables and cluster-level
variables are present. The cluster to which a unit belongs may then
influence the unit-level observations - and vice versa - for each of the
units that make up the cluster \citep{hox17}. These relations can and
should be taken into account when developing analysis models for
multilevel data for the simple reason that groups of observations share
some common variance.

@@

The process of analyzing multilevel data is further complicated when not
all observations are observed. Just as with single level data,
missingness may occur at the unit level. But with multiple levels of
data comes the potential for clustered missingness. Missingness in
multilevel data can therefore be categorized into two general patterns:
systematic missingness and sporadic missingness, see \citet{resc13}. We
have visualized the difference between these types of missingness in
Figure 1. The figure shows an \(n \times p\) set
\(\mathbf{X} = X_1, \dots, X_p\), with \(n=5\) units and \(p=3\) columns
distributed over 3 clusters. Column \texttt{X1} is completely observed.
Column \texttt{X2} is systematically missing and column \texttt{X3} is
sporadically missing.\footnote{Explain why. GV: not necessary}

@@

@@

Here we can set \texttt{maxit} to 1\ldots{}
